A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-23 06-appendix.tex
Teorema da terminação geral

\begin{definition}
  \label{def.Hnst}
  %%
  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
  true.  Nested-irreducible descriptions serve as normal forms for~$\nst$
  transitions: they embody the result of an exhaustive number of~$\nst$
  applications.  We will write~$\delta_\Hnst$ to indicate that
  description~$\delta$ is nested-irreducible.
\end{definition}

@@

\begin{lemma}\strut
  \begin{enumerate}[(a)]
  \item If~$\delta\outpush\delta'$ then~$\rank(\delta)=\rank(\delta')$.
  \item If~$\delta\outpop\delta'$ then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
    %%
    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}
    Therefore, $\<i,j>=\<i',j'>$.

@@

% \begin{theorem}[Reaction termination]
%   For any~$\delta$ there is a~$\delta'_\Hnst$ such
%   that~$\delta\trans[*]\delta'$ and~$\rank(\delta')=\<0,0>$.
% \end{theorem}
% \begin{proof}
%   By lexicographic induction on~$rank(\delta)$.
% \end{proof}
%   \begin{basis}
%     If~$\<i,j>=\<0,0>$ then, by~\ref{?}, $\delta$ cannot be advanced
%     by~$\out$ transitions.  There are two possibilities\dots\ (it can or
%     cannot be advanced by nst transitions).
%   \end{basis}
%   \begin{induction}
%     Let~$\<i,j>\ne\<0,0>$ and suppose that the theorem holds for
%     all~$\delta''$ such that~$\rank(\delta'')<\rank(\delta)$.
%     \begin{case}
%       $\delta$~is nested-irreducible.
%       %%
%       \begin{subcase}
%         $\delta\outpush\delta_1$, for some~$\delta_1$.
%       \end{subcase}
%       \begin{subcase}
%         $\delta\outpop\delta_1$, for some~$\delta_1$.
%         %%
%         Then, by definition of~\R{pop}, $\rank(\delta_1)=\<i,j-1>$, and by
%         the induction hypothesis\dots
%       \end{subcase}
%     \end{case}
%     \begin{case}
%       $\delta$~is not nested-irreducible.
%       %%
%     \end{case}
%   \end{induction}

>>>>

\begin{definition}
  \label{def.Hnst}
  %%
  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
  true.

  Nested-irreducible descriptions serve as normal forms for~$\nst$
  transitions: they embody the result of an exhaustive number of~$\nst$
  applications.  We will write~$\delta_\Hnst$ to indicate that
  description~$\delta$ is nested-irreducible.
\end{definition}

@@

\begin{lemma}\strut
  \label{lem.rank-out}
  \begin{enumerate}[(a)]
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
    %%
    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}
    Therefore, $\<i,j>=\<i',j'>$.

@@

\begin{theorem}
  \label{thm.rank-nst-*}
  %%
  If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{theorem}
\begin{proof}
  If~$\delta\nst[*]\delta'$ then there is an~$n$ such that
  $\delta\nst[n]\delta'$.  We proceed by induction on~$n$.
  %%
  The theorem is trivially true for~$n=0$ and follows directly from
  Lemma~\ref{lem.rank-nst} for~$n=1$.  Suppose
  $\delta\nst[1]\delta_1'\nst[n-1]\delta'$, for some~$n>1$ and~$\delta_1'$.
  Thus, by Lemma~\ref{lem.rank-nst} and by the induction hypothesis,
  \[
    \rank(\delta)\ge\rank(\delta_1')\ge\rank(\delta')\,.\qedhere
  \]
\end{proof}

\begin{definition}
  \label{def.H}
  %%
  A description~$\delta$ is \emph{irreducible} (in symbols, $\delta_\#$) iff
  it is nested-irreducible and its~$\rank(\delta)$ is~$\<i,0>$, for
  some~$i\ge0$.

  An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form
  for transitions in general.  It cannot be advanced by~$\nst$, as it is
  nested-irreducible, and neither by~$\outpush$ nor~$\outpop$, as the second
  coordinate of its rank is~0, which implies~$e=\nil$ and~$n=0$.
\end{definition}

\begin{theorem}
  \label{thm.term}
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  %%
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$
    transitions, as~$j=0$ implies~$e=\nil$ and~$n=0$.  Thus neither~\R{push}
    nor~\R{pop} can be applied.  There are two possibilities:
    either~$\delta$ is nested-irreducible or it is not.  In the first case,
    the theorem is trivially true, as~$\delta\nst[0]\delta_\Hnst$.

    Suppose~$\delta$ is not nested-irreducible.  Then, by
    Theorem~\ref{thm.term-nst-*}, $\delta\nst[*]\delta'_\Hnst$ for
    some~$\delta'_\Hnst$.  By Theorem~\ref{thm.rank-nst-*},
    \[
      \<i,j>=\<0,0>\ge\rank(\delta')\,,
    \]
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases.
    \begin{case}
      \label{thm.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      There are two cases.
      \begin{subcase}
        \label{thm.term.Hnst-j0}
        $j=0$.  By Definition~\ref{def.H}, $\delta_\#$.
        Thus~$\delta\trans[0]\delta_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.term.Hnst-j>0}
        $j>0$.
        %%
        There are two subcases.
        \begin{subsubcase}
          \label{thm.term.Hnst-j>0-nonnil}
          $e\ne\nil$.
          %%
          Then, by~\R{out} and by Theorem~\ref{thm.term-nst-*}, there
          are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
          \[
            \delta\outpush\delta_1'\nst[*]\delta'_\Hnst\,.
          \]
          Thus, by Lemma by item~\eqref{lem.rank-out-push} of
          Lemma~\ref{lem.rank-out} and by Theorem~\ref{thm.rank-nst-*},
          \begin{align*}
            \rank(\delta)=\rank(\delta_1')&=\<i,j>\\
                                          &\ge\rank(\delta')=\<i',j'>\,.
          \end{align*}
          If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof
          is similar to that of Case~\ref{thm.term.Hnst-j>0-nil}.
          Otherwise, if~$e'\ne\nil$ then $i>i'$, since
          an~$\ceu{\EmitInt(e')}$ was consumed by the nested transitions.
          Thus,
          \[
            \rank(\delta)>\rank(\delta')\,.
          \]
          By the induction hypothesis, $\delta'\trans[*]\delta''_\#$, for
          some~$\delta''_\#$.  Thus~$\delta\trans[*]\delta''_\#$.
        \end{subsubcase}
        \begin{subsubcase}
          \label{thm.term.Hnst-j>0-nil}
          $e=\nil$.
          %%
          Then, since~$j>0$, $\delta\outpop\delta'$, for
          some~$\delta''$.  By item~\eqref{lem.rank-out-pop} of
          Lemma~\ref{lem.rank-out},
          \[
            \rank(\delta)>\rank(\delta')\,.
          \]
          Hence, by the induction hypothesis, there is a~$\delta''_\#$ such
          that~$\delta'\trans[*]\delta''_\#$.
          Thus~$\delta\trans[*]\delta''_\#$.
        \end{subsubcase}
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      There are two subcases.
      \begin{subcase}
        $e\ne\nil$.  Similar to Case~\ref{thm.term.Hnst-j>0-nonnil}.
      \end{subcase}
      \begin{subcase}
        $e=\nil$.  Then, by Theorems~\ref{thm.rank-nst-*}
        and~\ref{thm.rank-nst-*} there is a~$\delta'_\Hnst$ such
        that~$\delta\nst[*]\delta'_\Hnst$
        with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
        is similar to that of Case~\ref{thm.term.Hnst}.\qedhere
      \end{subcase}
    \end{case}
  \end{induction}
\end{proof}
